---
title: "Gene Enrichment & GSEA Analysis Report"
subtitle: "Comprehensive Pathway and Ontology Enrichment Results"
author:
  - name: "Nicolas Blavet"
    affiliation: CEITEC-Central European Institute of Technology, Masaryk University
  - name: "Bioinformatics Core Facility"
    affiliation: Centre for Molecular Medicine, CEITEC Masaryk University
affiliations:
  - CEITEC-Central European Institute of Technology, Kamenice 5, 625 00 Brno, Czech Republic
output:
  bookdown::html_document2:
    theme: flatly
    highlight: tango
    number_sections: false
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    css: styles_modern.css
    self_contained: false
params:
  config: snakemake.params.config
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-packages, include=FALSE}
library(data.table)
library(jsonlite)
library(flextable)
library(stringr)
library(dplyr)

set_flextable_defaults(
  font.family = "Segoe UI",
  font.size = 11,
  padding.top = 4,
  padding.bottom = 4#,
  # border.top = NULL,
  # border.bottom = NULL,
  # border.left = NULL,
  # border.right = NULL
)
```

<style>
/* Custom CSS for modern look */
:root {
  --primary-color: #2c3e50;
  --secondary-color: #3498db;
  --accent-go: #9b59b6;
  --accent-kegg: #e74c3c;
  --accent-reactome: #f39c12;
  --accent-wp: #1abc9c;
  --accent-enrichr: #34495e;
  --bg-light: #f8f9fa;
  --card-shadow: 0 2px 8px rgba(0,0,0,0.1);
}

body {
  font-family: 'Segoe UI', Tahoma, Geneva, Verdana, sans-serif;
  line-height: 1.6;
  color: #333;
}

.container-main {
  max-width: 1400px;
  margin: 0 auto;
  padding: 20px;
}

/* Hero section */
.hero-section {
  background: linear-gradient(135deg, var(--primary-color) 0%, var(--secondary-color) 100%);
  color: white;
  padding: 40px;
  border-radius: 12px;
  margin-bottom: 30px;
  box-shadow: var(--card-shadow);
}

.hero-section h1 {
  color: white !important;
  margin-bottom: 10px;
  font-size: 2.2em;
}

.hero-section .subtitle {
  opacity: 0.9;
  font-size: 1.1em;
}

/* Card styling */
.result-card {
  background: white;
  border-radius: 10px;
  padding: 25px;
  margin-bottom: 25px;
  box-shadow: var(--card-shadow);
  border-left: 4px solid var(--secondary-color);
}

.result-card.go-card { border-left-color: var(--accent-go); }
.result-card.kegg-card { border-left-color: var(--accent-kegg); }
.result-card.reactome-card { border-left-color: var(--accent-reactome); }
.result-card.wp-card { border-left-color: var(--accent-wp); }
.result-card.enrichr-card { border-left-color: var(--accent-enrichr); }

/* Section headers */
.section-badge {
  display: inline-block;
  padding: 8px 16px;
  border-radius: 20px;
  color: white;
  font-weight: 600;
  margin-bottom: 15px;
}

.badge-go { background: var(--accent-go); }
.badge-kegg { background: var(--accent-kegg); }
.badge-reactome { background: var(--accent-reactome); }
.badge-wp { background: var(--accent-wp); }
.badge-enrichr { background: var(--accent-enrichr); }

/* Stat boxes */
.stat-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(200px, 1fr));
  gap: 20px;
  margin: 20px 0;
}

.stat-box {
  background: var(--bg-light);
  padding: 20px;
  border-radius: 8px;
  text-align: center;
  border: 1px solid #e0e0e0;
}

.stat-box .stat-value {
  font-size: 2em;
  font-weight: bold;
  color: var(--secondary-color);
}

.stat-box .stat-label {
  color: #666;
  font-size: 0.9em;
  margin-top: 5px;
}

/* Figure containers */
.figure-container {
  background: var(--bg-light);
  padding: 15px;
  border-radius: 8px;
  margin: 15px 0;
}

.figure-container img {
  max-width: 100%;
  height: auto;
  border-radius: 6px;
}

/* Table improvements */
.rmarkdown table {
  border-collapse: collapse;
  width: 100%;
  margin: 20px 0;
}

.rmarkdown th {
  background: var(--primary-color);
  color: white;
  padding: 12px;
  text-align: left;
}

.rmarkdown td {
  padding: 10px 12px;
  border-bottom: 1px solid #eee;
}

.rmarkdown tr:hover {
  background: #f5f5f5;
}

/* Alert/Info boxes */
.info-box {
  background: #e8f4fd;
  border-left: 4px solid var(--secondary-color);
  padding: 15px 20px;
  margin: 15px 0;
  border-radius: 0 8px 8px 0;
}

.warning-box {
  background: #fff8e1;
  border-left: 4px solid #ffc107;
  padding: 15px 20px;
  margin: 15px 0;
  border-radius: 0 8px 8px 0;
}

/* Summary dashboard */
.dashboard-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(280px, 1fr));
  gap: 20px;
  margin: 20px 0;
}

.dashboard-card {
  background: white;
  border-radius: 10px;
  padding: 20px;
  box-shadow: var(--card-shadow);
}

.dashboard-card h4 {
  margin-top: 0;
  color: var(--primary-color);
  border-bottom: 2px solid var(--secondary-color);
  padding-bottom: 10px;
}

/* Accordion-style sections */
details {
  margin: 10px 0;
  border: 1px solid #ddd;
  border-radius: 8px;
  overflow: hidden;
}

details summary {
  background: var(--bg-light);
  padding: 15px 20px;
  cursor: pointer;
  font-weight: 600;
  list-style: none;
  transition: background 0.2s;
}

details summary:hover {
  background: #e9ecef;
}

details[open] summary {
  background: var(--secondary-color);
  color: white;
}

details p {
  padding: 15px 20px;
  margin: 0;
}

/* Badge for significant results */
.sig-badge {
  display: inline-block;
  padding: 2px 8px;
  border-radius: 4px;
  font-size: 0.8em;
  font-weight: 600;
}

.sig-high { background: #d4edda; color: #155724; }
.sig-medium { background: #fff3cd; color: #856404; }
.sig-low { background: #f8d7da; color: #721c24; }

/* Code blocks */
pre {
  background: #2d2d2d;
  color: #f8f8f2;
  padding: 15px;
  border-radius: 6px;
  font-size: 0.9em;
}

/* Footer */
.report-footer {
  margin-top: 50px;
  padding: 30px;
  background: var(--bg-light);
  border-radius: 10px;
  text-align: center;
  color: #666;
}
</style>

```{r load-config-data, include=FALSE}
# Load config data - assumes running from enrichment_gsea/ directory
# where config_enrichment_gsea.json is located
config_json <- as.data.table(fromJSON(txt = params$config))
config_json[,condition := sapply(1:length(config_json$samples),function(x) config_json$samples[[x]]$condition)]
config_json[,tag       := sapply(1:length(config_json$samples),function(x) config_json$samples[[x]]$tag)]
config_json[,full_name := sapply(1:length(config_json$samples),function(x) config_json$samples[[x]]$sample_name)]

comparison <- unlist(strsplit(unique(config_json$comparison),split="|",fixed=TRUE))
biotype <- unique(config_json$biotype_list)
analysis_type <- unique(config_json$analysis_type)
analysis <- unlist(strsplit(analysis_type,split="|",fixed=TRUE))

DE <- paste0("DE_", analysis)

# Set species/organism display name
species <- if(!is.null(config_json$organism) && length(config_json$organism) > 0) {
  tools::toTitleCase(gsub("_", " ", config_json$organism[1]))
} else {
  "Unknown organism"
}

# Helper: build a table of result files (DE_analysis/comparison/subdir/file.tsv)
# with parsed analysis, comparison and enrich columns
# Same approach as the original report: for each DE analysis, list the result
# file recursively inside the DE_* folder (the filename is unique enough)
build_result_table <- function(subdir, filename){
  dt <- data.table(tablpath = character(0))
  for(d in DE){
    files <- list.files(path = d, pattern = gsub(".", "\\.", filename),
                        full.names = TRUE, recursive = TRUE)
    if(length(files) > 0){
      dt <- rbindlist(list(dt, data.table(tablpath = files)), fill = TRUE)
    }
  }
  if(nrow(dt) == 0){
    return(data.table(tablpath = character(0), de = character(0), comp = character(0), enrich = character(0)))
  }
  dt[, de := unlist(strsplit(tablpath, "/"))[1], by = .(tablpath)]
  dt[, comp := unlist(strsplit(tablpath, "/"))[2], by = .(tablpath)]
  # enrich is the suffix of the subdirectory name (e.g. enrichment_GO_all -> all)
  dt[, subdir_name := unlist(strsplit(tablpath, "/"))[3], by = .(tablpath)]
  dt[, enrich := ifelse(grepl("_(all|up|down)$", subdir_name), sub(".*_", "", subdir_name), "")]
  dt
}

# Helper: print a figure container with a readable title
# (plot name - analysis - comparison [- gene set])
print_figure <- function(plotpath, de, comp, enrich, alt_text, plot_title = "Enrichment"){
  title <- paste0(plot_title, " - ", de, " - ", comp)
  if(!is.na(enrich) && length(enrich) > 0 && nzchar(enrich)){
    title <- paste0(title, " (", enrich, ")")
  }
  cat('<div class="figure-container">')
  cat('<h4>', title, '</h4>')
  if(file.exists(plotpath)){
    cat('<img src="', plotpath, '" alt="', alt_text, '">', sep="")
  }
  cat('</div>')
}

# Helper: display plots from the summary_enrich_stats table (built in the
# summary-dashboard chunk). Rows are selected by prefix on the Enrichment
# column (e.g. "enrichment_KEGG" matches enrichment_KEGG_all/up/down).
# Enriched == 0 -> warning message instead of the plot.
# Title = Analysis - Comparison - Enrichment
show_result_files <- function(stats, pattern, alt_text, warn_msg){
  dt <- stats[grepl(paste0("^", pattern), Enrichment)]
  if(nrow(dt) == 0){
    cat('<div class="info-box">No results available for this database.</div>')
    return(invisible())
  }
  for(i in seq_len(nrow(dt))){
    title <- paste0(dt$Analysis[i], " - ", dt$Comparison[i], " - ", dt$Enrichment[i])
    if(dt$Enriched[i] == 0){
      cat(paste0('<div class="warning-box"><strong>', warn_msg, '</strong> for ', title, '</div>'))
    } else {
      cat('<div class="figure-container">')
      cat('<h4>', title, '</h4>')
      if(file.exists(dt$plot_path[i])){
        cat('<img src="', dt$plot_path[i], '" alt="', alt_text, '">', sep="")
      }
      cat('</div>')
    }
  }
}
```

<div class="container-main">

```{r hero-section, echo=FALSE, results='asis'}
cat('<div class="hero-section">')
  cat('<h1>🧬 Gene Enrichment & GSEA Analysis</h1>')
  cat('<p class="subtitle">Comprehensive Pathway and Ontology Enrichment Results</p>')
  cat('<div style="margin-top: 20px;">')
    cat(paste0('<p style="color: white; margin: 5px 0;"><strong>Generated:</strong> ', format(Sys.time(), "%B %d, %Y") ,' </p>'))
    cat(paste0('<p style="color: white; margin: 5px 0;"><strong>Organism:</strong> ', species, '</p>'))
  cat('</div>')
cat('</div>')
```

<!-- Summary Dashboard -->
```{r summary-dashboard, echo=FALSE, results='asis'}
# Calculate per-comparison summary statistics
summary_gene_stats <- data.table()
for (d in DE){
  for(f in comparison){
    de_path <- paste0(d, "/", f, "/gene_for_gsea.tsv")
    if(file.exists(de_path)){
      de <- fread(de_path, header = T)
      names(de)[1] <- 'Geneid'

      n_gsea <- length(de$Geneid)
      n_all <- length(de[abs(log2FoldChange) >= as.numeric(config_json$cutoff_log2fc_enrich[1]) &
         padj <= as.numeric(config_json$cutoff_padj_enrich[1]), ]$Geneid)
      n_up <- length(de[log2FoldChange >= as.numeric(config_json$cutoff_log2fc_enrich[1]) &
         padj <= as.numeric(config_json$cutoff_padj_enrich[1]), ]$Geneid)
      n_down <- length(de[log2FoldChange <= -as.numeric(config_json$cutoff_log2fc_enrich[1]) &
         padj <= as.numeric(config_json$cutoff_padj_enrich[1]), ]$Geneid)

      dt <- data.table(Analysis=d, Comparison=f, `Genes for enrichment (all)`=n_all,
                        `Genes for enrichment (up)`=n_up, `Genes for enrichment (down)`=n_down,
                        `Genes for GSEA`=n_gsea)
      
      summary_gene_stats <- rbind(summary_gene_stats, dt)
    }
  }
}

go_type <- c("BP","MF","CC")
enrich_type <- c("all","up","down")

summary_enrich_stats <- data.table()
for (d in DE){
  for(f in comparison){
    if(config_json$onthology[1] == T){
      for(h in go_type){
        for(g in enrich_type){
          enr_path <- paste0(d, "/", f, "/enrichment_GO_", g, "/enrich_GO_", h, ".tsv")
          if(file.exists(enr_path)){
          enr <- fread(enr_path, header = T)
          n_enr <- length(enr$ID)
          dtt <- data.table(Analysis=d, Comparison=f, Enrichment=paste0("enrichment_GO_", h,"_",g), Enriched=n_enr, plot_path=gsub("tsv$","svg",enr_path))
          summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
          }
        }

        gsea_path <- paste0(d, "/", f, "/GSEA_GO/GSEA_GO_", h, ".tsv")
        if(file.exists(gsea_path)){
          gsea_file <- fread(gsea_path, header = T)
          n_gsea <- length(gsea_file$ID)
          dtt <- data.table(Analysis=d, Comparison=f, Enrichment=paste0("GSEA_GO_", h), Enriched=n_gsea, plot_path=gsub("tsv$","svg",gsea_path))
          summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
        }
      }
    }
    if(config_json$kegg[1] == T){
      for(g in enrich_type){
        enr_path <- paste0(d, "/", f, "/enrichment_KEGG_", g, "/enrich_KEGG.tsv")
        if(file.exists(enr_path)){
          enr <- fread(enr_path, header = T)
          n_enr <- length(enr$ID)
          dtt <- data.table(Analysis=d, Comparison=f, Enrichment=paste0("enrichment_KEGG_",g), Enriched=n_enr, plot_path=gsub("tsv$","svg",enr_path))
          summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
        }
      }
      gsea_path <- paste0(d, "/", f, "/GSEA_KEGG/GSEA_KEGG.tsv")
      if(file.exists(gsea_path)){
        gsea_file <- fread(gsea_path, header = T)
        n_gsea <- length(gsea_file$ID)
        dtt <- data.table(Analysis=d, Comparison=f, Enrichment="GSEA_KEGG", Enriched=n_gsea, plot_path=gsub("tsv$","svg",gsea_path))
        summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
      }
    }
    if(config_json$reactome[1] == T){
      for(g in enrich_type){
        enr_path <- paste0(d, "/", f, "/enrichment_REACTOME_", g, "/enrich_REACTOME.tsv")
        if(file.exists(enr_path)){
          enr <- fread(enr_path, header = T)
          n_enr <- length(enr$ID)
          dtt <- data.table(Analysis=d, Comparison=f, Enrichment=paste0("enrichment_REACTOME_",g), Enriched=n_enr, plot_path=gsub("tsv$","svg",enr_path))
          summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
        }
      }
      gsea_path <- paste0(d, "/", f, "/GSEA_REACTOME/GSEA_REACTOME.tsv")
      if(file.exists(gsea_path)){
        gsea_file <- fread(gsea_path, header = T)
        n_gsea <- length(gsea_file$ID)
        dtt <- data.table(Analysis=d, Comparison=f, Enrichment="GSEA_REACTOME", Enriched=n_gsea, plot_path=gsub("tsv$","svg",gsea_path))
        summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
      }
    }
    if(config_json$wikipathways[1] == T){
      for(g in enrich_type){
        enr_path <- paste0(d, "/", f, "/enrichment_WP_", g, "/enrich_WP.tsv")
        if(file.exists(enr_path)){
          enr <- fread(enr_path, header = T)
          n_enr <- length(enr$ID)
          dtt <- data.table(Analysis=d, Comparison=f, Enrichment=paste0("enrichment_WP_",g), Enriched=n_enr, plot_path=gsub("tsv$","svg",enr_path))
          summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
        }
      }
      gsea_path <- paste0(d, "/", f, "/GSEA_WP/GSEA_WP.tsv")
      if(file.exists(gsea_path)){
        gsea_file <- fread(gsea_path, header = T)
        n_gsea <- length(gsea_file$ID)
        dtt <- data.table(Analysis=d, Comparison=f, Enrichment="GSEA_WP", Enriched=n_gsea, plot_path=gsub("tsv$","svg",gsea_path))
        summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
      }
    }
    if(config_json$gseapy_enrichr[1] == T){
      for(g in enrich_type){
        enr_path <- paste0(d, "/", f, "/enrichr_", g, "/", config_json$enrichR_DB[1], "." , config_json$organism_enrichR[1], ".enrichr.reports.txt")
        if(file.exists(enr_path)){
          enr <- fread(enr_path, header = T)
          n_enr <- length(enr$Term)
          dtt <- data.table(Analysis=d, Comparison=f, Enrichment=paste0("enrichR-",config_json$enrichR_DB[1],"_",g), Enriched=n_enr, plot_path=gsub("txt$","svg",enr_path))
          summary_enrich_stats <- rbind(summary_enrich_stats, dtt)
        }
      }
    }
  }
}

# Render dashboard - one cat per line to avoid pandoc issues
cat('<div class="dashboard-grid">')
cat('<div class="dashboard-card">')
cat('<h4>Experimental Design</h4>')
cat('<div class="stat-grid">')
cat(paste0('<div class="stat-box"><div class="stat-value">', length(config_json[,full_name]), '</div><div class="stat-label">Total Samples</div></div>'))
cat(paste0('<div class="stat-box"><div class="stat-value">', length(config_json[,unique(condition)]), '</div><div class="stat-label">Conditions</div></div>'))
cat(paste0('<div class="stat-box"><div class="stat-value">', length(comparison), '</div><div class="stat-label">Comparisons</div></div>'))
cat('</div>')
cat('</div>')
cat('<div class="dashboard-card">')
cat('<h4>Analysis Parameters</h4>')
cat('<div class="stat-grid">')
cat(paste0('<div class="stat-box"><div class="stat-value">±', config_json$cutoff_log2fc_enrich[1], '</div><div class="stat-label">Enrichment Log2FC</div></div>'))
cat(paste0('<div class="stat-box"><div class="stat-value">', config_json$cutoff_padj_enrich[1], '</div><div class="stat-label">Enrichment Padj</div></div>'))
cat(paste0('<div class="stat-box"><div class="stat-value">', config_json$n_up[1], '</div><div class="stat-label">Top Terms Displayed</div></div>'))
cat('</div>')
cat('</div>')
cat('</div>')
```

```{r summary-tables-display, echo=FALSE, results='asis'}
# Plain HTML table printer (flextable output is unreliable inside cat()/asis chunks)
cat_sticky_table <- function(dt, max_height = 380){
  cat('<style>')
  cat('.scroll-table{max-height:', max_height, 'px;overflow-y:auto;border:1px solid #e0e0e0;border-radius:4px;margin-top:10px;}', sep = "")
  cat('.scroll-table table{width:100%;border-collapse:collapse;margin:0;}')
  cat('.scroll-table th{position:sticky;top:0;z-index:2;background:#2c3e50;color:white;padding:8px 10px;text-align:center;border:none;}')
  cat('.scroll-table td{padding:6px 10px;text-align:center;border:none;border-bottom:1px solid #eee;background:white;}')
  cat('.scroll-table tr:nth-child(even) td{background:#f8f9fa;}')
  cat('</style>')
  cat('<div class="scroll-table">')
  cat('<table>')
  cat('<thead><tr>', paste0('<th>', colnames(dt), '</th>', collapse = ""), '</tr></thead>')
  cat('<tbody>')
  for(i in seq_len(nrow(dt))){
    cat('<tr>', paste0('<td>', as.character(unlist(dt[i, ])), '</td>', collapse = ""), '</tr>')
  }
  cat('</tbody>')
  cat('</table>')
  cat('</div>')
}

cat('<div class="result-card">')
cat('<h3 style="margin-top: 0;">Genes per Analysis and Comparison</h3>')
cat(paste0('<p>Number of genes passing the enrichment cutoffs (|log2FC| ≥ ',config_json$cutoff_log2fc_enrich[1],' padj ≤ ', config_json$cutoff_padj_enrich[1], ') and used for GSEA:</p>'))
if(nrow(summary_gene_stats) > 0){
  cat_sticky_table(summary_gene_stats)
}
cat('</div>')

cat('<div class="result-card">')
cat('<h3 style="margin-top: 0;">Enriched Terms per Analysis and Comparison</h3>')
cat(paste0('<p>Number of significant terms (adjusted p-value ≤ ',config_json$enrich_padj[1],') found by each enrichment/GSEA method:</p>'))
if(nrow(summary_enrich_stats) > 0){
  cat_sticky_table(summary_enrich_stats[, !c("plot_path"), with = FALSE])
}
cat('</div>')
```


<div class="info-box">
<strong>About this report:</strong> This report presents gene enrichment and Gene Set Enrichment Analysis (GSEA) results.
Enrichment analysis identifies over-represented functional terms among differentially expressed genes,
while GSEA detects subtle but coordinated changes across predefined gene sets.
</div>

<!-- Acknowledgement Section -->
<div class="result-card">
<h2>Acknowledgements</h2>
<ul>
<li><strong>Co-authorship request:</strong> If we have developed novel tools, algorithms, or pipelines, participated in experiment design planning, or contributed to a biological question addressed in your work, we request co-authorship. We will assist with manuscript preparation.</li>
<li><strong>Core Facility acknowledgment:</strong> Core Facility Bioinformatics of CEITEC Masaryk University is gratefully acknowledged for the obtaining of the scientific data presented in this analysis.</li>
</ul>
</div>

<!-- Description Section -->
<div class="result-card">
<h2>Analysis Overview</h2>

```{r description-section, echo=FALSE, results='asis'}
# Use organism field instead of species (which may not exist in config)
species <- if(!is.null(config_json$organism) && length(config_json$organism) > 0) {
  tools::toTitleCase(gsub("_", " ", config_json$organism[1]))
} else {
  "Unknown organism"
}

n_repl <- round(length(config_json[,full_name])/length(config_json[,unique(condition)]))
cat('<div class="info-box">')
cat('<p>This analysis processes differential expression results from DESeq2 and performs pathway enrichment analysis across multiple databases.</p>')
cat('<ul>')
cat(paste0('<li><strong>Organism:</strong> ', species, '</li>'))
cat(paste0('<li><strong>Total samples:</strong> ', length(config_json[,full_name]), '</li>'))
cat(paste0('<li><strong>Conditions:</strong> ', paste(unique(config_json[,condition]), collapse=", "), '</li>'))
cat(paste0('<li><strong>Replicates per condition:</strong> ~', n_repl, '</li>'))
cat('</ul>')
cat('</div>')

# Comparisons table
comparison_dt <- data.table()
for (d in DE){
  for(f in comparison){
    de_path <- paste0(d, "/", f, "/DESeq2.tsv")
    if(file.exists(de_path)){
      de <- fread(de_path, header = T)
      names(de)[1] <- 'Geneid'

      deseq_cutoff <- function(deseq, LOG2FC, PADJ){
        x <- deseq[is.na(padj) == F & is.na(pvalue) == F,]
        x <- x[abs(log2FoldChange) >= LOG2FC & padj <= PADJ,]
        return(x)
      }

      deseq2_enrich <- deseq_cutoff(de, as.numeric(config_json$cutoff_log2fc_enrich[1]), as.numeric(config_json$cutoff_padj_enrich[1]))
      deseq2_gsea <- deseq_cutoff(de, as.numeric(config_json$cutoff_log2fc_gsea[1]), as.numeric(config_json$cutoff_padj_gsea[1]))

      dt <- data.table(
        Analysis = d,
        Comparison = f,
        "Genes for Enrichment" = length(deseq2_enrich$Geneid),
        "Genes for GSEA" = length(deseq2_gsea$Geneid),
        "Total Genes" = length(de$Geneid)
      )
      comparison_dt <- rbind(comparison_dt, dt)
    }
  }
}

# Create styled comparison table
if(nrow(comparison_dt) > 0){
  cat('<details open>')
  cat('<summary>Comparison Summary Table (click to collapse)</summary>')
  cat('<p>Number of genes considered for each analysis type across all comparisons:</p>')

  ft <- qflextable(data = as.data.frame(comparison_dt))
  ft <- ft %>%
    theme_booktabs() %>%
    set_header_labels(
      Analysis = "Analysis Type",
      Comparison = "Comparison",
      "Genes for Enrichment" = "Enrichment Genes",
      "Genes for GSEA" = "GSEA Genes",
      "Total Genes" = "Total"
    ) %>%
    align(align = "center") %>%
    color(part = "header", color = "white") %>%
    bg(part = "header", bg = "#2c3e50")

  print(ft)
  cat('</details>')
}
```

<h3 style="margin-top: 30px;">Agreed Analysis Tasks</h3>
<div class="stat-grid">
`r if(config_json$onthology[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #9b59b6;"><div class="stat-value" style="color:#9b59b6;">GO</div><div class="stat-label">Gene Ontology<br><small>BP / MF / CC</small></div></div>')`
`r if(config_json$kegg[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #e74c3c;"><div class="stat-value" style="color:#e74c3c;">KEGG</div><div class="stat-label">Pathways<br><small>Metabolic & Signaling</small></div></div>')`
`r if(config_json$reactome[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #f39c12;"><div class="stat-value" style="color:#f39c12;">Reactome</div><div class="stat-label">Pathways<br><small>Biological Processes</small></div></div>')`
`r if(config_json$wikipathways[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #1abc9c;"><div class="stat-value" style="color:#1abc9c;">WikiPathways</div><div class="stat-label">Pathways<br><small>Community-curated</small></div></div>')`
`r if(config_json$gseapy_enrichr[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #34495e;"><div class="stat-value" style="color:#34495e;">EnrichR</div><div class="stat-label">', config_json$enrichR_DB[1], '<br><small>Library</small></div></div>')`
</div>

<h3 style="margin-top: 30px;">Methodology</h3>
<div class="info-box">
<strong>Gene Enrichment Analysis (ORA):</strong>
<p>Based on genes identified as Differentially Expressed with |log2FoldChange| ≥ `r config_json$cutoff_log2fc_enrich[1]` and adjusted p-value ≤ `r config_json$cutoff_padj_enrich[1]`. Tests whether specific functional terms are over-represented among DE genes using hypergeometric test.</p>
</div>

<div class="info-box">
<strong>Gene Set Enrichment Analysis (GSEA):</strong>
<p>Considers all genes with valid p-values from DESeq2, ranked by log2FoldChange. Detects coordinated subtle changes in predefined gene sets without arbitrary cutoffs. Uses fgsea algorithm for efficient permutation-based significance estimation.</p>
</div>

<div class="info-box">
<strong>Statistical Parameters:</strong>
<ul>
<li><strong>p-adjustment method:</strong> Benjamini-Hochberg (BH)</li>
<li><strong>Enrichment p-value cutoff:</strong> `r config_json$enrich_padj[1]`</li>
<li><strong>GSEA permutations:</strong> 10,000</li>
<li><strong>Gene set size range:</strong> `r config_json$enrich_minGSSize[1]` - `r config_json$enrich_maxGSSize[1]` genes</li>
</ul>
</div>
</div>

<!-- GO Section -->
`r if(config_json$onthology[1]==TRUE) '<div class=\"result-card go-card\">'`
`r if(config_json$onthology[1]==TRUE) '<span class=\"section-badge badge-go\">Gene Ontology</span>'`
`r if(config_json$onthology[1]==TRUE) '<h2>Gene Ontology Analysis</h2>'`

`r if(config_json$onthology[1]==TRUE) '<h3>Biological Process (BP)</h3>'`
`r if(config_json$onthology[1]==TRUE) '<details><summary>Enrichment Analysis</summary>'`
```{r enrich-BP-modern, echo=FALSE, results='asis'}
if(config_json$onthology[1]==TRUE){
  show_result_files(summary_enrich_stats, "enrichment_GO_BP", "GO BP Enrichment", "No significant enrichment")
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<details><summary>GSEA Analysis</summary>'`
```{r gsea-BP-modern, echo=FALSE, results='asis'}
if(config_json$onthology[1]==TRUE){
  show_result_files(summary_enrich_stats, "GSEA_GO_BP", "GSEA GO BP", "No gene set enrichment")
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<h3>Molecular Function (MF)</h3>'`
`r if(config_json$onthology[1]==TRUE) '<details><summary>Enrichment Analysis</summary>'`
```{r enrich-MF-modern, echo=FALSE, results='asis'}
if(config_json$onthology[1]==TRUE){
  show_result_files(summary_enrich_stats, "enrichment_GO_MF", "GO MF Enrichment", "No significant enrichment")
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<details><summary>GSEA Analysis</summary>'`
```{r gsea-MF-modern, echo=FALSE, results='asis'}
if(config_json$onthology[1]==TRUE){
  show_result_files(summary_enrich_stats, "GSEA_GO_MF", "GSEA GO MF", "No gene set enrichment")
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<h3>Cellular Component (CC)</h3>'`
`r if(config_json$onthology[1]==TRUE) '<details><summary>Enrichment Analysis</summary>'`
```{r enrich-CC-modern, echo=FALSE, results='asis'}
if(config_json$onthology[1]==TRUE){
  show_result_files(summary_enrich_stats, "enrichment_GO_CC", "GO CC Enrichment", "No significant enrichment")
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<details><summary>GSEA Analysis</summary>'`
```{r gsea-CC-modern, echo=FALSE, results='asis'}
if(config_json$onthology[1]==TRUE){
  show_result_files(summary_enrich_stats, "GSEA_GO_CC", "GSEA GO CC", "No gene set enrichment")
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '</div>'`

<!-- KEGG Section -->
`r if(config_json$kegg[1]==TRUE) '<div class=\"result-card kegg-card\">'`
`r if(config_json$kegg[1]==TRUE) '<span class=\"section-badge badge-kegg\">KEGG Pathways</span>'`
`r if(config_json$kegg[1]==TRUE) '<h2>KEGG Pathway Analysis</h2>'`

`r if(config_json$kegg[1]==TRUE) '<details><summary>Enrichment Analysis</summary>'`
```{r enrich-kegg-modern, echo=FALSE, results='asis'}
if(config_json$kegg[1]==TRUE){
  show_result_files(summary_enrich_stats, "enrichment_KEGG", "KEGG Enrichment", "No significant enrichment")
}
```
`r if(config_json$kegg[1]==TRUE) '</details>'`

`r if(config_json$kegg[1]==TRUE) '<details><summary>GSEA Analysis</summary>'`
```{r gsea-kegg-modern, echo=FALSE, results='asis'}
if(config_json$kegg[1]==TRUE){
  show_result_files(summary_enrich_stats, "GSEA_KEGG", "GSEA KEGG", "No gene set enrichment")
}
```
`r if(config_json$kegg[1]==TRUE) '</details>'`
`r if(config_json$kegg[1]==TRUE) '</div>'`

<!-- Reactome Section -->
`r if(config_json$reactome[1]==TRUE) '<div class=\"result-card reactome-card\">'`
`r if(config_json$reactome[1]==TRUE) '<span class=\"section-badge badge-reactome\">Reactome Pathways</span>'`
`r if(config_json$reactome[1]==TRUE) '<h2>Reactome Pathway Analysis</h2>'`

`r if(config_json$reactome[1]==TRUE) '<details><summary>Enrichment Analysis</summary>'`
```{r enrich-reactome-modern, echo=FALSE, results='asis'}
if(config_json$reactome[1]==TRUE){
  show_result_files(summary_enrich_stats, "enrichment_REACTOME", "Reactome Enrichment", "No significant enrichment")
}
```
`r if(config_json$reactome[1]==TRUE) '</details>'`

`r if(config_json$reactome[1]==TRUE) '<details><summary>GSEA Analysis</summary>'`
```{r gsea-reactome-modern, echo=FALSE, results='asis'}
if(config_json$reactome[1]==TRUE){
  show_result_files(summary_enrich_stats, "GSEA_REACTOME", "GSEA Reactome", "No gene set enrichment")
}
```
`r if(config_json$reactome[1]==TRUE) '</details>'`
`r if(config_json$reactome[1]==TRUE) '</div>'`

<!-- WikiPathways Section -->
`r if(config_json$wikipathways[1]==TRUE) '<div class=\"result-card wp-card\">'`
`r if(config_json$wikipathways[1]==TRUE) '<span class=\"section-badge badge-wp\">WikiPathways</span>'`
`r if(config_json$wikipathways[1]==TRUE) '<h2>WikiPathways Analysis</h2>'`

`r if(config_json$wikipathways[1]==TRUE) '<details><summary>Enrichment Analysis</summary>'`
```{r enrich-wp-modern, echo=FALSE, results='asis'}
if(config_json$wikipathways[1]==TRUE){
  show_result_files(summary_enrich_stats, "enrichment_WP", "WikiPathways Enrichment", "No significant enrichment")
}
```
`r if(config_json$wikipathways[1]==TRUE) '</details>'`

`r if(config_json$wikipathways[1]==TRUE) '<details><summary>GSEA Analysis</summary>'`
```{r gsea-wp-modern, echo=FALSE, results='asis'}
if(config_json$wikipathways[1]==TRUE){
  show_result_files(summary_enrich_stats, "GSEA_WP", "GSEA WikiPathways", "No gene set enrichment")
}
```
`r if(config_json$wikipathways[1]==TRUE) '</details>'`
`r if(config_json$wikipathways[1]==TRUE) '</div>'`

<!-- EnrichR Section -->
`r if(config_json$gseapy_enrichr[1]==TRUE) '<div class=\"result-card enrichr-card\">'`
`r if(config_json$gseapy_enrichr[1]==TRUE) '<span class=\"section-badge badge-enrichr\">EnrichR</span>'`
`r if(config_json$gseapy_enrichr[1]==TRUE) '<h2>EnrichR Analysis</h2>'`

`r if(config_json$gseapy_enrichr[1]==TRUE) paste0('<details><summary>EnrichR Results (', config_json$enrichR_DB[1], ')</summary>')`
```{r enrich-enrichr-modern, echo=FALSE, results='asis'}
if(config_json$gseapy_enrichr[1]==TRUE){
  show_result_files(summary_enrich_stats, "enrichR", "EnrichR", "No gene set enrichment")
}
```
`r if(config_json$gseapy_enrichr[1]==TRUE) '</details>'`
`r if(config_json$gseapy_enrichr[1]==TRUE) '</div>'`

<!-- Output Files Section -->
<div class="result-card">
<h2>Output Files Reference</h2>

<div class="info-box">
<strong>Directory Structure:</strong><br>
Results are organized by <code>Analysis/Comparison/</code>. All files in a folder share the same comparison context.
</div>

<h3>Main Input Files</h3>
<table class="file-table">
<tr><th>File</th><th>Description</th></tr>
<tr><td><code>gene_for_enrichment.tsv</code></td><td>DE genes used for enrichment analysis (filtered by log2FC and padj cutoffs)</td></tr>
<tr><td><code>gene_for_gsea.tsv</code></td><td>All genes with valid p-values, ranked by log2FC for GSEA</td></tr>
</table>

<h3>Enrichment Output Files</h3>
`r if(config_json$onthology[1]==TRUE) '<p><strong>Gene Ontology:</strong> <code>enrichment_GO_{all|up|down}/enrich_GO_{BP|MF|CC}.{tsv|svg|pdf}</code></p>'`
`r if(config_json$kegg[1]==TRUE) '<p><strong>KEGG:</strong> <code>enrichment_KEGG_{all|up|down}/enrich_KEGG.{tsv|svg|pdf}</code></p>'`
`r if(config_json$reactome[1]==TRUE) '<p><strong>Reactome:</strong> <code>enrichment_REACTOME_{all|up|down}/enrich_REACTOME.{tsv|svg|pdf}</code></p>'`
`r if(config_json$wikipathways[1]==TRUE) '<p><strong>WikiPathways:</strong> <code>enrichment_WP_{all|up|down}/enrich_WP.{tsv|svg|pdf}</code></p>'`
`r if(config_json$gseapy_enrichr[1]==TRUE) paste0('<p><strong>EnrichR:</strong> <code>enrichr_{all|up|down}/', config_json$enrichR_DB[1], '.', config_json$organism_enrichR[1], '.enrichr.reports.{txt|svg}</code></p>')`

<h3>GSEA Output Files</h3>
`r if(config_json$onthology[1]==TRUE) '<p><strong>Gene Ontology:</strong> <code>GSEA_GO/GSEA_GO_{BP|MF|CC}.{tsv|svg|pdf}</code></p>'`
`r if(config_json$kegg[1]==TRUE) '<p><strong>KEGG:</strong> <code>GSEA_KEGG/GSEA_KEGG.{tsv|svg|pdf}</code></p>'`
`r if(config_json$reactome[1]==TRUE) '<p><strong>Reactome:</strong> <code>GSEA_REACTOME/GSEA_REACTOME.{tsv|svg|pdf}</code></p>'`
`r if(config_json$wikipathways[1]==TRUE) '<p><strong>WikiPathways:</strong> <code>GSEA_WP/GSEA_WP.{tsv|svg|pdf}</code></p>'`

<h3>Column Definitions</h3>

<div class="details">
<summary><strong>Enrichment Result Columns</strong></summary>
<ul>
<li><strong>ID</strong> - Feature ID (ontology term or pathway identifier)</li>
<li><strong>Description</strong> - Functional description of the term/pathway</li>
<li><strong>GeneRatio</strong> - K/L format: K genes from input set have this annotation / L genes from input set have any annotation</li>
<li><strong>BgRatio</strong> - M/N format: M genes in database have this annotation / N total genes in database</li>
<li><strong>pvalue</strong> - Raw enrichment p-value (hypergeometric test)</li>
<li><strong>p.adjust</strong> - Benjamini-Hochberg adjusted p-value (use this for significance!)</li>
<li><strong>qvalue</strong> - False discovery rate estimate</li>
<li><strong>geneID</strong> - List of genes contributing to this enrichment</li>
<li><strong>Count</strong> - Number of genes sharing this feature</li>
</ul>
</div>

<div class="details">
<summary><strong>GSEA Result Columns</strong></summary>
<ul>
<li><strong>ID</strong> - Feature ID</li>
<li><strong>Description</strong> - Functional description</li>
<li><strong>setSize</strong> - Number of genes in the gene set</li>
<li><strong>enrichmentScore</strong> - Running sum maximum deviation (ES)</li>
<li><strong>NES</strong> - Normalized Enrichment Score (comparable across datasets)</li>
<li><strong>pvalue</strong> - Raw GSEA p-value</li>
<li><strong>p.adjust</strong> - FDR-adjusted p-value</li>
<li><strong>rank</strong> - Position in ranked list where ES occurs</li>
<li><strong>leading_edge</strong> - Core genes driving enrichment (Tags/List/Signal)</li>
<li><strong>core_enrichment</strong> - List of leading edge genes</li>
</ul>
</div>
</div>

<!-- Footer -->
<div class="report-footer">
<p><strong>Bioinformatics Core Facility</strong> | Centre for Molecular Medicine, CEITEC Masaryk University</p>
<p>For questions or custom analysis requests, contact: nicolas.blavet@ceitec.muni.cz</p>
<p style="font-size: 0.9em; color: #999;">Report generated with the enrichment_and_gsea pipeline | `r format(Sys.time(), "%Y-%m-%d %H:%M")`</p>
</div>

</div> <!-- End container-main -->
