---
title: "Gene Enrichment & GSEA Analysis Report"
subtitle: "Comprehensive Pathway and Ontology Enrichment Results"
author:
  - name: "Nicolas Blavet"
    affiliation: CEITEC-Central European Institute of Technology, Masaryk University
  - name: "Bioinformatics Core Facility"
    affiliation: Centre for Molecular Medicine, CEITEC Masaryk University
affiliations:
  - CEITEC-Central European Institute of Technology, Kamenice 5, 625 00 Brno, Czech Republic
output:
  bookdown::html_document2:
    theme: flatly
    highlight: tango
    number_sections: false
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    css: styles_modern.css
    self_contained: false
params:
  # config parameter kept for snakemake compatibility but not used
  config: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load-packages, include=FALSE}
library(data.table)
library(jsonlite)
library(flextable)
library(stringr)
library(dplyr)

set_flextable_defaults(
  font.family = "Segoe UI",
  font.size = 11,
  padding.top = 4,
  padding.bottom = 4,
  border.top = NULL,
  border.bottom = NULL,
  border.left = NULL,
  border.right = NULL
)
```

```{r variables, results='asis', echo=FALSE}
`%!in%` <- Negate(`%in%`)
```

<style>
/* Custom CSS for modern look */
:root {
  --primary-color: #2c3e50;
  --secondary-color: #3498db;
  --accent-go: #9b59b6;
  --accent-kegg: #e74c3c;
  --accent-reactome: #f39c12;
  --accent-wp: #1abc9c;
  --accent-enrichr: #34495e;
  --bg-light: #f8f9fa;
  --card-shadow: 0 2px 8px rgba(0,0,0,0.1);
}

body {
  font-family: 'Segoe UI', Tahoma, Geneva, Verdana, sans-serif;
  line-height: 1.6;
  color: #333;
}

.container-main {
  max-width: 1400px;
  margin: 0 auto;
  padding: 20px;
}

/* Hero section */
.hero-section {
  background: linear-gradient(135deg, var(--primary-color) 0%, var(--secondary-color) 100%);
  color: white;
  padding: 40px;
  border-radius: 12px;
  margin-bottom: 30px;
  box-shadow: var(--card-shadow);
}

.hero-section h1 {
  color: white !important;
  margin-bottom: 10px;
  font-size: 2.2em;
}

.hero-section .subtitle {
  opacity: 0.9;
  font-size: 1.1em;
}

/* Card styling */
.result-card {
  background: white;
  border-radius: 10px;
  padding: 25px;
  margin-bottom: 25px;
  box-shadow: var(--card-shadow);
  border-left: 4px solid var(--secondary-color);
}

.result-card.go-card { border-left-color: var(--accent-go); }
.result-card.kegg-card { border-left-color: var(--accent-kegg); }
.result-card.reactome-card { border-left-color: var(--accent-reactome); }
.result-card.wp-card { border-left-color: var(--accent-wp); }
.result-card.enrichr-card { border-left-color: var(--accent-enrichr); }

/* Section headers */
.section-badge {
  display: inline-block;
  padding: 8px 16px;
  border-radius: 20px;
  color: white;
  font-weight: 600;
  margin-bottom: 15px;
}

.badge-go { background: var(--accent-go); }
.badge-kegg { background: var(--accent-kegg); }
.badge-reactome { background: var(--accent-reactome); }
.badge-wp { background: var(--accent-wp); }
.badge-enrichr { background: var(--accent-enrichr); }

/* Stat boxes */
.stat-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(200px, 1fr));
  gap: 20px;
  margin: 20px 0;
}

.stat-box {
  background: var(--bg-light);
  padding: 20px;
  border-radius: 8px;
  text-align: center;
  border: 1px solid #e0e0e0;
}

.stat-box .stat-value {
  font-size: 2em;
  font-weight: bold;
  color: var(--secondary-color);
}

.stat-box .stat-label {
  color: #666;
  font-size: 0.9em;
  margin-top: 5px;
}

/* Figure containers */
.figure-container {
  background: var(--bg-light);
  padding: 15px;
  border-radius: 8px;
  margin: 15px 0;
}

.figure-container img {
  max-width: 100%;
  height: auto;
  border-radius: 6px;
}

/* Table improvements */
.rmarkdown table {
  border-collapse: collapse;
  width: 100%;
  margin: 20px 0;
}

.rmarkdown th {
  background: var(--primary-color);
  color: white;
  padding: 12px;
  text-align: left;
}

.rmarkdown td {
  padding: 10px 12px;
  border-bottom: 1px solid #eee;
}

.rmarkdown tr:hover {
  background: #f5f5f5;
}

/* Alert/Info boxes */
.info-box {
  background: #e8f4fd;
  border-left: 4px solid var(--secondary-color);
  padding: 15px 20px;
  margin: 15px 0;
  border-radius: 0 8px 8px 0;
}

.warning-box {
  background: #fff8e1;
  border-left: 4px solid #ffc107;
  padding: 15px 20px;
  margin: 15px 0;
  border-radius: 0 8px 8px 0;
}

/* Summary dashboard */
.dashboard-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(280px, 1fr));
  gap: 20px;
  margin: 20px 0;
}

.dashboard-card {
  background: white;
  border-radius: 10px;
  padding: 20px;
  box-shadow: var(--card-shadow);
}

.dashboard-card h4 {
  margin-top: 0;
  color: var(--primary-color);
  border-bottom: 2px solid var(--secondary-color);
  padding-bottom: 10px;
}

/* Accordion-style sections */
details {
  margin: 10px 0;
  border: 1px solid #ddd;
  border-radius: 8px;
  overflow: hidden;
}

details summary {
  background: var(--bg-light);
  padding: 15px 20px;
  cursor: pointer;
  font-weight: 600;
  list-style: none;
  transition: background 0.2s;
}

details summary:hover {
  background: #e9ecef;
}

details[open] summary {
  background: var(--secondary-color);
  color: white;
}

details p {
  padding: 15px 20px;
  margin: 0;
}

/* Badge for significant results */
.sig-badge {
  display: inline-block;
  padding: 2px 8px;
  border-radius: 4px;
  font-size: 0.8em;
  font-weight: 600;
}

.sig-high { background: #d4edda; color: #155724; }
.sig-medium { background: #fff3cd; color: #856404; }
.sig-low { background: #f8d7da; color: #721c24; }

/* Code blocks */
pre {
  background: #2d2d2d;
  color: #f8f8f2;
  padding: 15px;
  border-radius: 6px;
  font-size: 0.9em;
}

/* Footer */
.report-footer {
  margin-top: 50px;
  padding: 30px;
  background: var(--bg-light);
  border-radius: 10px;
  text-align: center;
  color: #666;
}
</style>

```{r load-config-data, include=FALSE}
# Load config data - assumes running from enrichment_gsea/ directory
# where config_enrichment_gsea.json is located
config_json <- as.data.table(fromJSON("config_enrichment_gsea.json"))
config_json[,condition := sapply(1:length(config_json$samples),function(x) config_json$samples[[x]]$condition)]
config_json[,tag := sapply(1:length(config_json$samples),function(x) config_json$samples[[x]]$tag)]
config_json[,full_name := sapply(1:length(config_json$samples),function(x) config_json$samples[[x]]$sample_name)]

comparison <- unlist(strsplit(unique(config_json$comparison),split="|",fixed=TRUE))
biotype <- unique(config_json$biotype_list)
analysis_type <- unique(config_json$analysis_type)
analysis <- unlist(strsplit(analysis_type,split="|",fixed=TRUE))

DE <- paste0("DE_", analysis)

# Set species/organism display name
species <- if(!is.null(config_json$organism) && length(config_json$organism) > 0) {
  tools::toTitleCase(gsub("_", " ", config_json$organism[1]))
} else {
  "Unknown organism"
}
```

<div class="container-main">

<!-- Hero Header -->
<div class="hero-section">
  <h1>🧬 Gene Enrichment & GSEA Analysis</h1>
  <p class="subtitle">Comprehensive Pathway and Ontology Enrichment Results</p>
  <div style="margin-top: 20px; opacity: 0.9;">
    <p><strong>Generated:</strong> `r format(Sys.time(), "%B %d, %Y at %H:%M")`</p>
    <p><strong>Organism:</strong> `r species`</p>
  </div>
</div>

<!-- Summary Dashboard -->
```{r summary-dashboard, echo=FALSE, results='asis'}
# Calculate summary statistics
summary_stats <- data.table()
for (d in DE){
  for(f in comparison){
    de_path <- paste0(d, "/", f, "/DESeq2.tsv")
    if(file.exists(de_path)){
      de <- fread(de_path, header = T)
      names(de)[1] <- 'Geneid'

      deseq_cutoff <- function(deseq, LOG2FC, PADJ){
        x <- deseq[is.na(padj) == F & is.na(pvalue) == F,]
        x <- x[abs(log2FoldChange) >= LOG2FC & padj <= PADJ,]
        return(x)
      }

      enrich_genes <- deseq_cutoff(de, as.numeric(config_json$cutoff_log2fc_enrich[1]), as.numeric(config_json$cutoff_padj_enrich[1]))
      gsea_genes <- deseq_cutoff(de, as.numeric(config_json$cutoff_log2fc_gsea[1]), as.numeric(config_json$cutoff_padj_gsea[1]))

      # Count significant enrichments
      n_enrich <- 0
      databases <- c()
      if(config_json$onthology[1]==TRUE){
        bp_file <- paste0("../", d, "/", f, "/enrichment_GO_all/enrich_GO_BP.tsv")
        if(file.exists(bp_file)){
          bp_res <- fread(bp_file)
          if(nrow(bp_res) > 0) n_enrich <- n_enrich + nrow(bp_res)
        }
        databases <- c(databases, "GO")
      }
      if(config_json$kegg[1]==TRUE){
        kegg_file <- paste0("../", d, "/", f, "/enrich_KEGG_all/enrich_KEGG.tsv")
        if(file.exists(kegg_file)){
          kegg_res <- fread(kegg_file)
          if(nrow(kegg_res) > 0) n_enrich <- n_enrich + nrow(kegg_res)
        }
        databases <- c(databases, "KEGG")
      }

      dt <- data.table(
        Analysis = d,
        Comparison = f,
        Total_Genes = length(de$Geneid),
        Enrich_Genes = length(enrich_genes$Geneid),
        GSEA_Genes = length(gsea_genes$Geneid),
        Significant_Terms = n_enrich,
        Databases = paste(unique(databases), collapse=", ")
      )
      summary_stats <- rbind(summary_stats, dt)
    }
  }
}

# Render dashboard
cat('
<div class="dashboard-grid">
  <div class="dashboard-card">
    <h4>📊 Experimental Design</h4>
')
cat(paste0('    <div class="stat-grid">
      <div class="stat-box">
        <div class="stat-value">', length(config_json[,full_name]), '</div>
        <div class="stat-label">Total Samples</div>
      </div>
      <div class="stat-box">
        <div class="stat-value">', length(config_json[,unique(condition)]), '</div>
        <div class="stat-label">Conditions</div>
      </div>
      <div class="stat-box">
        <div class="stat-value">', length(comparison), '</div>
        <div class="stat-label">Comparisons</div>
      </div>
    </div>
'))
cat('
  </div>

  <div class="dashboard-card">
    <h4>🔬 Analysis Parameters</h4>
')
cat(paste0('    <div class="stat-grid">
      <div class="stat-box">
        <div class="stat-value">±', config_json$cutoff_log2fc_enrich[1], '</div>
        <div class="stat-label">Enrichment Log2FC</div>
      </div>
      <div class="stat-box">
        <div class="stat-value">', config_json$cutoff_padj_enrich[1], '</div>
        <div class="stat-label">Enrichment Padj</div>
      </div>
      <div class="stat-box">
        <div class="stat-value">', config_json$n_up[1], '</div>
        <div class="stat-label">Top Terms Displayed</div>
      </div>
    </div>
'))
cat('
  </div>

  <div class="dashboard-card">
    <h4>📈 Summary Statistics</h4>
')
total_significant <- sum(summary_stats$Significant_Terms, na.rm=TRUE)
avg_enrich_genes <- round(mean(summary_stats$Enrich_Genes, na.rm=TRUE))
cat(paste0('    <div class="stat-grid">
      <div class="stat-box">
        <div class="stat-value">', total_significant, '</div>
        <div class="stat-label">Significant Terms</div>
      </div>
      <div class="stat-box">
        <div class="stat-value">', avg_enrich_genes, '</div>
        <div class="stat-label">Avg DE Genes</div>
      </div>
      <div class="stat-box">
        <div class="stat-value">', nrow(summary_stats), '</div>
        <div class="stat-label">Analysis Sets</div>
      </div>
    </div>
'))
cat('
  </div>
</div>
')
```

<div class="info-box">
<strong>ℹ️ About this report:</strong> This report presents gene enrichment and Gene Set Enrichment Analysis (GSEA) results.
Enrichment analysis identifies over-represented functional terms among differentially expressed genes,
while GSEA detects subtle but coordinated changes across predefined gene sets.
</div>

<!-- Acknowledgement Section -->
<div class="result-card">
<h2>🙏 Acknowledgements</h2>
<ul>
<li><strong>Co-authorship request:</strong> If we have developed novel tools, algorithms, or pipelines, participated in experiment design planning, or contributed to a biological question addressed in your work, we request co-authorship. We will assist with manuscript preparation.</li>
<li><strong>Core Facility acknowledgment:</strong> Core Facility Bioinformatics of CEITEC Masaryk University is gratefully acknowledged for the obtaining of the scientific data presented in this analysis.</li>
</ul>
</div>

<!-- Description Section -->
<div class="result-card">
<h2>📋 Analysis Overview</h2>

```{r description-section, echo=FALSE, results='asis'}
# Use organism field instead of species (which may not exist in config)
species <- if(!is.null(config_json$organism) && length(config_json$organism) > 0) {
  tools::toTitleCase(gsub("_", " ", config_json$organism[1]))
} else {
  "Unknown organism"
}

cat(paste0('
<div class="info-box">
<p>This analysis processes differential expression results from DESeq2 and performs pathway enrichment analysis across multiple databases.</p>
<ul>
<li><strong>Organism:</strong> ', species, '</li>
<li><strong>Total samples:</strong> ', length(config_json[,full_name]), '</li>
<li><strong>Conditions:</strong> ', paste(unique(config_json[,condition]), collapse=", "), '</li>
<li><strong>Replicates per condition:</strong> ~', round(length(config_json[,full_name])/length(config_json[,unique(condition)])), '</li>
</ul>
</div>
'))

# Comparisons table
comparison_dt <- data.table()
for (d in DE){
  for(f in comparison){
    de_path <- paste0(d, "/", f, "/DESeq2.tsv")
    if(file.exists(de_path)){
      de <- fread(de_path, header = T)
      names(de)[1] <- 'Geneid'

      deseq_cutoff <- function(deseq, LOG2FC, PADJ){
        x <- deseq[is.na(padj) == F & is.na(pvalue) == F,]
        x <- x[abs(log2FoldChange) >= LOG2FC & padj <= PADJ,]
        return(x)
      }

      deseq2_enrich <- deseq_cutoff(de, as.numeric(config_json$cutoff_log2fc_enrich[1]), as.numeric(config_json$cutoff_padj_enrich[1]))
      deseq2_gsea <- deseq_cutoff(de, as.numeric(config_json$cutoff_log2fc_gsea[1]), as.numeric(config_json$cutoff_padj_gsea[1]))

      dt <- data.table(
        Analysis = d,
        Comparison = f,
        "Genes for Enrichment" = length(deseq2_enrich$Geneid),
        "Genes for GSEA" = length(deseq2_gsea$Geneid),
        "Total Genes" = length(de$Geneid)
      )
      comparison_dt <- rbind(comparison_dt, dt)
    }
  }
}

# Create styled comparison table
if(nrow(comparison_dt) > 0){
  cat('<details open>')
  cat('<summary>📊 Comparison Summary Table (click to collapse)</summary>')
  cat('<p>Number of genes considered for each analysis type across all comparisons:</p>')

  ft <- qflextable(data = as.data.frame(comparison_dt))
  ft <- ft %>%
    theme_booktabs() %>%
    set_header_labels(
      Analysis = "Analysis Type",
      Comparison = "Comparison",
      "Genes for Enrichment" = "Enrichment Genes",
      "Genes for GSEA" = "GSEA Genes",
      "Total Genes" = "Total"
    ) %>%
    align(align = "center") %>%
    header_style(color = "white", bg = "#2c3e50")

  print(ft)
  cat('</details>')
}
```

<h3 style="margin-top: 30px;">🎯 Agreed Analysis Tasks</h3>
<div class="stat-grid">
`r if(config_json$onthology[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #9b59b6;"><div class="stat-value" style="color:#9b59b6;">GO</div><div class="stat-label">Gene Ontology<br><small>BP / MF / CC</small></div></div>')`
`r if(config_json$kegg[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #e74c3c;"><div class="stat-value" style="color:#e74c3c;">KEGG</div><div class="stat-label">Pathways<br><small>Metabolic & Signaling</small></div></div>')`
`r if(config_json$reactome[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #f39c12;"><div class="stat-value" style="color:#f39c12;">Reactome</div><div class="stat-label">Pathways<br><small>Biological Processes</small></div></div>')`
`r if(config_json$wikipathways[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #1abc9c;"><div class="stat-value" style="color:#1abc9c;">WikiPathways</div><div class="stat-label">Pathways<br><small>Community-curated</small></div></div>')`
`r if(config_json$gseapy_enrichr[1]==TRUE) paste0('<div class="stat-box" style="border-left: 4px solid #34495e;"><div class="stat-value" style="color:#34495e;">EnrichR</div><div class="stat-label">', config_json$enrichR_DB[1], '<br><small>Library</small></div></div>')`
</div>

<h3 style="margin-top: 30px;">📖 Methodology</h3>
<div class="info-box">
<strong>Gene Enrichment Analysis (ORA):</strong>
<p>Based on genes identified as Differentially Expressed with |log2FoldChange| ≥ `r config_json$cutoff_log2fc_enrich[1]` and adjusted p-value ≤ `r config_json$cutoff_padj_enrich[1]`. Tests whether specific functional terms are over-represented among DE genes using hypergeometric test.</p>
</div>

<div class="info-box">
<strong>Gene Set Enrichment Analysis (GSEA):</strong>
<p>Considers all genes with valid p-values from DESeq2, ranked by log2FoldChange. Detects coordinated subtle changes in predefined gene sets without arbitrary cutoffs. Uses fgsea algorithm for efficient permutation-based significance estimation.</p>
</div>

<div class="info-box">
<strong>Statistical Parameters:</strong>
<ul>
<li><strong>p-adjustment method:</strong> Benjamini-Hochberg (BH)</li>
<li><strong>Enrichment p-value cutoff:</strong> `r config_json$enrich_padj[1]`</li>
<li><strong>GSEA permutations:</strong> 10,000</li>
<li><strong>Gene set size range:</strong> `r config_json$enrich_minGSSize[1]` - `r config_json$enrich_maxGSSize[1]` genes</li>
</ul>
</div>
</div>

<!-- GO Section -->
`r if(config_json$onthology[1]==TRUE) '<div class=\"result-card go-card\">'`
`r if(config_json$onthology[1]==TRUE) '<span class=\"section-badge badge-go\">Gene Ontology</span>'`
`r if(config_json$onthology[1]==TRUE) '<h2>🧬 Gene Ontology Analysis</h2>'`

`r if(config_json$onthology[1]==TRUE) '<details><summary>📌 Biological Process (BP)</summary>'`
```{r enrich-BP-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$onthology[1]==TRUE){
  BP.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "enrich_GO_BP.tsv", full.names = T, recursive = T))
  if(nrow(BP.files) == 0) BP.files <- data.table(tablpath = list.files(path = ".", pattern = "enrich_GO_BP.tsv", full.names = T, recursive = T))

  BP.files <- BP.files[, plotpath := gsub(".tsv",".svg",tablpath)]
  BP.files <- BP.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(BP.files) > 0){
    for(i in 1:nrow(BP.files)){
      dt <- tryCatch(fread(BP.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No significant enrichment</strong> found in ', BP.files$de[i], ' - ', BP.files$comp[i], ' for GO Biological Process.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', BP.files$de[i], ' | ', BP.files$comp[i], '</h4>')
        if(file.exists(BP.files$plotpath[i])){
          cat('<img src="', BP.files$plotpath[i], '" alt="GO BP Enrichment">', sep="")
        }
        cat('</div>')
      }
    }
  } else {
    cat('<div class="info-box">No Biological Process enrichment plots available.</div>')
  }
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<details><summary>📌 Molecular Function (MF)</summary>'`
```{r enrich-MF-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$onthology[1]==TRUE){
  MF.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "enrich_GO_MF.tsv", full.names = T, recursive = T))
  if(nrow(MF.files) == 0) MF.files <- data.table(tablpath = list.files(path = ".", pattern = "enrich_GO_MF.tsv", full.names = T, recursive = T))

  MF.files <- MF.files[, plotpath := gsub(".tsv",".svg",tablpath)]
  MF.files <- MF.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(MF.files) > 0){
    for(i in 1:nrow(MF.files)){
      dt <- tryCatch(fread(MF.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No significant enrichment</strong> found in ', MF.files$de[i], ' - ', MF.files$comp[i], ' for GO Molecular Function.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', MF.files$de[i], ' | ', MF.files$comp[i], '</h4>')
        if(file.exists(MF.files$plotpath[i])){
          cat('<img src="', MF.files$plotpath[i], '" alt="GO MF Enrichment">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<details><summary>📌 Cellular Component (CC)</summary>'`
```{r enrich-CC-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$onthology[1]==TRUE){
  CC.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "enrich_GO_CC.tsv", full.names = T, recursive = T))
  if(nrow(CC.files) == 0) CC.files <- data.table(tablpath = list.files(path = ".", pattern = "enrich_GO_CC.tsv", full.names = T, recursive = T))

  CC.files <- CC.files[, plotpath := gsub(".tsv",".svg",tablpath)]
  CC.files <- CC.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(CC.files) > 0){
    for(i in 1:nrow(CC.files)){
      dt <- tryCatch(fread(CC.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No significant enrichment</strong> found in ', CC.files$de[i], ' - ', CC.files$comp[i], ' for GO Cellular Component.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', CC.files$de[i], ' | ', CC.files$comp[i], '</h4>')
        if(file.exists(CC.files$plotpath[i])){
          cat('<img src="', CC.files$plotpath[i], '" alt="GO CC Enrichment">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '<hr>'`
`r if(config_json$onthology[1]==TRUE) '<h3 style="color:#9b59b6;">🔄 GSEA - Gene Set Enrichment Analysis</h3>'`

`r if(config_json$onthology[1]==TRUE) '<details><summary>📊 GSEA Biological Process</summary>'`
```{r gsea-BP-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$onthology[1]==TRUE){
  BPg.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "GSEA_GO_BP.tsv", full.names = TRUE, recursive = TRUE))
  if(nrow(BPg.files) == 0) BPg.files <- data.table(tablpath = list.files(path = ".", pattern = "GSEA_GO_BP.tsv", full.names = TRUE, recursive = TRUE))

  BPg.files <- BPg.files[, plotpath := gsub(".tsv", ".svg", tablpath)]
  BPg.files <- BPg.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(BPg.files) > 0){
    for(i in 1:nrow(BPg.files)){
      dt <- tryCatch(fread(BPg.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No gene set enrichment</strong> detected in ', BPg.files$de[i], ' - ', BPg.files$comp[i], ' for GO Biological Process.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', BPg.files$de[i], ' | ', BPg.files$comp[i], '</h4>')
        if(file.exists(BPg.files$plotpath[i])){
          cat('<img src="', BPg.files$plotpath[i], '" alt="GSEA GO BP">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$onthology[1]==TRUE) '</details>'`

`r if(config_json$onthology[1]==TRUE) '</div>'`

<!-- KEGG Section -->
`r if(config_json$kegg[1]==TRUE) '<div class=\"result-card kegg-card\">'`
`r if(config_json$kegg[1]==TRUE) '<span class=\"section-badge badge-kegg\">KEGG Pathways</span>'`
`r if(config_json$kegg[1]==TRUE) '<h2>🔬 KEGG Pathway Analysis</h2>'`

`r if(config_json$kegg[1]==TRUE) '<details open><summary>📍 Enrichment Analysis</summary>'`
```{r enrich-kegg-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$kegg[1]==TRUE){
  ke.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "enrich_KEGG.tsv", full.names = TRUE, recursive = TRUE))
  if(nrow(ke.files) == 0) ke.files <- data.table(tablpath = list.files(path = ".", pattern = "enrich_KEGG.tsv", full.names = TRUE, recursive = TRUE))

  ke.files <- ke.files[, plotpath := gsub(".tsv", ".svg", tablpath)]
  ke.files <- ke.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(ke.files) > 0){
    for(i in 1:nrow(ke.files)){
      dt <- tryCatch(fread(ke.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No significant enrichment</strong> found in ', ke.files$de[i], ' - ', ke.files$comp[i], ' for KEGG pathways.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', ke.files$de[i], ' | ', ke.files$comp[i], '</h4>')
        if(file.exists(ke.files$plotpath[i])){
          cat('<img src="', ke.files$plotpath[i], '" alt="KEGG Enrichment">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$kegg[1]==TRUE) '</details>'`

`r if(config_json$kegg[1]==TRUE) '<details><summary>📍 GSEA Analysis</summary>'`
```{r gsea-kegg-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$kegg[1]==TRUE){
  keg.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "GSEA_KEGG.tsv", full.names = TRUE, recursive = TRUE))
  if(nrow(keg.files) == 0) keg.files <- data.table(tablpath = list.files(path = ".", pattern = "GSEA_KEGG.tsv", full.names = TRUE, recursive = TRUE))

  keg.files <- keg.files[, plotpath := gsub(".tsv", ".svg", tablpath)]
  keg.files <- keg.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(keg.files) > 0){
    for(i in 1:nrow(keg.files)){
      dt <- tryCatch(fread(keg.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No gene set enrichment</strong> detected in ', keg.files$de[i], ' - ', keg.files$comp[i], ' for KEGG pathways.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', keg.files$de[i], ' | ', keg.files$comp[i], '</h4>')
        if(file.exists(keg.files$plotpath[i])){
          cat('<img src="', keg.files$plotpath[i], '" alt="GSEA KEGG">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$kegg[1]==TRUE) '</details>'`
`r if(config_json$kegg[1]==TRUE) '</div>'`

<!-- Reactome Section -->
`r if(config_json$reactome[1]==TRUE) '<div class=\"result-card reactome-card\">'`
`r if(config_json$reactome[1]==TRUE) '<span class=\"section-badge badge-reactome\">Reactome Pathways</span>'`
`r if(config_json$reactome[1]==TRUE) '<h2>🧪 Reactome Pathway Analysis</h2>'`

`r if(config_json$reactome[1]==TRUE) '<details open><summary>📍 Enrichment Analysis</summary>'`
```{r enrich-reactome-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$reactome[1]==TRUE){
  re.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "enrich_REACTOME.tsv", full.names = TRUE, recursive = TRUE))
  if(nrow(re.files) == 0) re.files <- data.table(tablpath = list.files(path = ".", pattern = "enrich_REACTOME.tsv", full.names = TRUE, recursive = TRUE))

  re.files <- re.files[, plotpath := gsub(".tsv", ".svg", tablpath)]
  re.files <- re.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(re.files) > 0){
    for(i in 1:nrow(re.files)){
      dt <- tryCatch(fread(re.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No significant enrichment</strong> found in ', re.files$de[i], ' - ', re.files$comp[i], ' for Reactome pathways.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', re.files$de[i], ' | ', re.files$comp[i], '</h4>')
        if(file.exists(re.files$plotpath[i])){
          cat('<img src="', re.files$plotpath[i], '" alt="Reactome Enrichment">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$reactome[1]==TRUE) '</details>'`

`r if(config_json$reactome[1]==TRUE) '<details><summary>📍 GSEA Analysis</summary>'`
```{r gsea-reactome-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$reactome[1]==TRUE){
  reg.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "GSEA_REACTOME.tsv", full.names = TRUE, recursive = TRUE))
  if(nrow(reg.files) == 0) reg.files <- data.table(tablpath = list.files(path = ".", pattern = "GSEA_REACTOME.tsv", full.names = TRUE, recursive = TRUE))

  reg.files <- reg.files[, plotpath := gsub(".tsv", ".svg", tablpath)]
  reg.files <- reg.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(reg.files) > 0){
    for(i in 1:nrow(reg.files)){
      dt <- tryCatch(fread(reg.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No gene set enrichment</strong> detected in ', reg.files$de[i], ' - ', reg.files$comp[i], ' for Reactome pathways.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', reg.files$de[i], ' | ', reg.files$comp[i], '</h4>')
        if(file.exists(reg.files$plotpath[i])){
          cat('<img src="', reg.files$plotpath[i], '" alt="GSEA Reactome">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$reactome[1]==TRUE) '</details>'`
`r if(config_json$reactome[1]==TRUE) '</div>'`

<!-- WikiPathways Section -->
`r if(config_json$wikipathways[1]==TRUE) '<div class=\"result-card wp-card\">'`
`r if(config_json$wikipathways[1]==TRUE) '<span class=\"section-badge badge-wp\">WikiPathways</span>'`
`r if(config_json$wikipathways[1]==TRUE) '<h2>📚 WikiPathways Analysis</h2>'`

`r if(config_json$wikipathways[1]==TRUE) '<details open><summary>📍 Enrichment Analysis</summary>'`
```{r enrich-wp-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$wikipathways[1]==TRUE){
  wp.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "enrich_WP.tsv", full.names = TRUE, recursive = TRUE))
  if(nrow(wp.files) == 0) wp.files <- data.table(tablpath = list.files(path = ".", pattern = "enrich_WP.tsv", full.names = TRUE, recursive = TRUE))

  wp.files <- wp.files[, plotpath := gsub(".tsv", ".svg", tablpath)]
  wp.files <- wp.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(wp.files) > 0){
    for(i in 1:nrow(wp.files)){
      dt <- tryCatch(fread(wp.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No significant enrichment</strong> found in ', wp.files$de[i], ' - ', wp.files$comp[i], ' for WikiPathways.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', wp.files$de[i], ' | ', wp.files$comp[i], '</h4>')
        if(file.exists(wp.files$plotpath[i])){
          cat('<img src="', wp.files$plotpath[i], '" alt="WikiPathways Enrichment">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$wikipathways[1]==TRUE) '</details>'`

`r if(config_json$wikipathways[1]==TRUE) '<details><summary>📍 GSEA Analysis</summary>'`
```{r gsea-wp-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$wikipathways[1]==TRUE){
  wpg.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "GSEA_WP.tsv", full.names = TRUE, recursive = TRUE))
  if(nrow(wpg.files) == 0) wpg.files <- data.table(tablpath = list.files(path = ".", pattern = "GSEA_WP.tsv", full.names = TRUE, recursive = TRUE))

  wpg.files <- wpg.files[, plotpath := gsub(".tsv", ".svg", tablpath)]
  wpg.files <- wpg.files[, `:=` (de = unlist(strsplit(tablpath, "/"))[1], comp = unlist(strsplit(tablpath, "/"))[2]), by = .(tablpath)]

  if(nrow(wpg.files) > 0){
    for(i in 1:nrow(wpg.files)){
      dt <- tryCatch(fread(wpg.files$tablpath[i]), error = function(e) data.table())
      if(nrow(dt) == 0 || length(dt$ID) == 0){
        cat('<div class="warning-box"><strong>⚠️ No gene set enrichment</strong> detected in ', wpg.files$de[i], ' - ', wpg.files$comp[i], ' for WikiPathways.</div>')
      } else {
        cat('<div class="figure-container">')
        cat('<h4>', wpg.files$de[i], ' | ', wpg.files$comp[i], '</h4>')
        if(file.exists(wpg.files$plotpath[i])){
          cat('<img src="', wpg.files$plotpath[i], '" alt="GSEA WikiPathways">', sep="")
        }
        cat('</div>')
      }
    }
  }
}
```
`r if(config_json$wikipathways[1]==TRUE) '</details>'`
`r if(config_json$wikipathways[1]==TRUE) '</div>'`

<!-- EnrichR Section -->
`r if(config_json$gseapy_enrichr[1]==TRUE) '<div class=\"result-card enrichr-card\">'`
`r if(config_json$gseapy_enrichr[1]==TRUE) '<span class=\"section-badge badge-enrichr\">EnrichR</span>'`
`r if(config_json$gseapy_enrichr[1]==TRUE) '<h2>🔍 EnrichR Analysis</h2>'`

`r if(config_json$gseapy_enrichr[1]==TRUE) '<details open><summary>📍 EnrichR Results (' , config_json$enrichR_DB[1], ')</summary>'`
```{r enrich-enrichr-modern, echo=FALSE, results='asis', fig.width=10, fig.height=6, out.width="100%", dev="svg"}
if(config_json$gseapy_enrichr[1]==TRUE){
  enr.files <- data.table(tablpath = list.files(path = "DE_*", pattern = "enrichr_all/", full.names = TRUE, recursive = TRUE))
  if(nrow(enr.files) == 0) {
    enr.files <- data.table(tablpath = list.files(path = ".", pattern = "KEGG_2021_Human.human.enrichr.reports.svg", full.names = TRUE, recursive = TRUE))
  }

  # Try to find enrichr plots
  enr_plots <- data.table(
    plotpath = list.files(path = ".", pattern = "enrichr.*/.*\\.svg$", full.names = TRUE, recursive = TRUE)
  )

  if(nrow(enr_plots) > 0){
    for(i in 1:min(nrow(enr_plots), 10)){  # Limit to first 10 plots
      cat('<div class="figure-container">')
      cat('<h4>EnrichR - ', config_json$enrichR_DB[1], '</h4>')
      cat('<img src="', enr_plots$plotpath[i], '" alt="EnrichR Results">', sep="")
      cat('</div>')
    }
  } else {
    cat('<div class="info-box">EnrichR results will appear here when available.</div>')
  }
}
```
`r if(config_json$gseapy_enrichr[1]==TRUE) '</details>'`
`r if(config_json$gseapy_enrichr[1]==TRUE) '</div>'`

<!-- Output Files Section -->
<div class="result-card">
<h2>📁 Output Files Reference</h2>

<div class="info-box">
<strong>Directory Structure:</strong><br>
Results are organized by <code>Analysis/Comparison/</code>. All files in a folder share the same comparison context.
</div>

<h3>📊 Main Input Files</h3>
<table class="file-table">
<tr><th>File</th><th>Description</th></tr>
<tr><td><code>gene_for_enrichment.tsv</code></td><td>DE genes used for enrichment analysis (filtered by log2FC and padj cutoffs)</td></tr>
<tr><td><code>gene_for_gsea.tsv</code></td><td>All genes with valid p-values, ranked by log2FC for GSEA</td></tr>
</table>

<h3>📈 Enrichment Output Files</h3>
`r if(config_json$onthology[1]==TRUE) '<p><strong>Gene Ontology:</strong> <code>enrichment_GO_{all|up|down}/enrich_GO_{BP|MF|CC}.{tsv|svg|pdf}</code></p>'`
`r if(config_json$kegg[1]==TRUE) '<p><strong>KEGG:</strong> <code>enrichment_KEGG_{all|up|down}/enrich_KEGG.{tsv|svg|pdf}</code></p>'`
`r if(config_json$reactome[1]==TRUE) '<p><strong>Reactome:</strong> <code>enrichment_REACTOME_{all|up|down}/enrich_REACTOME.{tsv|svg|pdf}</code></p>'`
`r if(config_json$wikipathways[1]==TRUE) '<p><strong>WikiPathways:</strong> <code>enrichment_WP_{all|up|down}/enrich_WP.{tsv|svg|pdf}</code></p>'`
`r if(config_json$gseapy_enrichr[1]==TRUE) paste0('<p><strong>EnrichR:</strong> <code>enrichr_{all|up|down}/', config_json$enrichR_DB[1], '.', config_json$organism_enrichR[1], '.enrichr.reports.{txt|svg}</code></p>')`

<h3>🔄 GSEA Output Files</h3>
`r if(config_json$onthology[1]==TRUE) '<p><strong>Gene Ontology:</strong> <code>GSEA_GO_{all|up|down}/GSEA_GO_{BP|MF|CC}.{tsv|svg|pdf}</code></p>'`
`r if(config_json$kegg[1]==TRUE) '<p><strong>KEGG:</strong> <code>GSEA_KEGG_{all|up|down}/GSEA_KEGG.{tsv|svg|pdf}</code></p>'`
`r if(config_json$reactome[1]==TRUE) '<p><strong>Reactome:</strong> <code>GSEA_REACTOME_{all|up|down}/GSEA_REACTOME.{tsv|svg|pdf}</code></p>'`
`r if(config_json$wikipathways[1]==TRUE) '<p><strong>WikiPathways:</strong> <code>GSEA_WP_{all|up|down}/GSEA_WP.{tsv|svg|pdf}</code></p>'`

<h3>📋 Column Definitions</h3>

<div class="details">
<summary><strong>Enrichment Result Columns</strong></summary>
<ul>
<li><strong>ID</strong> - Feature ID (ontology term or pathway identifier)</li>
<li><strong>Description</strong> - Functional description of the term/pathway</li>
<li><strong>GeneRatio</strong> - K/L format: K genes from input set have this annotation / L genes from input set have any annotation</li>
<li><strong>BgRatio</strong> - M/N format: M genes in database have this annotation / N total genes in database</li>
<li><strong>pvalue</strong> - Raw enrichment p-value (hypergeometric test)</li>
<li><strong>p.adjust</strong> - Benjamini-Hochberg adjusted p-value (use this for significance!)</li>
<li><strong>qvalue</strong> - False discovery rate estimate</li>
<li><strong>geneID</strong> - List of genes contributing to this enrichment</li>
<li><strong>Count</strong> - Number of genes sharing this feature</li>
</ul>
</div>

<div class="details">
<summary><strong>GSEA Result Columns</strong></summary>
<ul>
<li><strong>ID</strong> - Feature ID</li>
<li><strong>Description</strong> - Functional description</li>
<li><strong>setSize</strong> - Number of genes in the gene set</li>
<li><strong>enrichmentScore</strong> - Running sum maximum deviation (ES)</li>
<li><strong>NES</strong> - Normalized Enrichment Score (comparable across datasets)</li>
<li><strong>pvalue</strong> - Raw GSEA p-value</li>
<li><strong>p.adjust</strong> - FDR-adjusted p-value</li>
<li><strong>rank</strong> - Position in ranked list where ES occurs</li>
<li><strong>leading_edge</strong> - Core genes driving enrichment (Tags/List/Signal)</li>
<li><strong>core_enrichment</strong> - List of leading edge genes</li>
</ul>
</div>
</div>

<!-- Footer -->
<div class="report-footer">
<p><strong>Bioinformatics Core Facility</strong> | Centre for Molecular Medicine, CEITEC Masaryk University</p>
<p>For questions or custom analysis requests, contact: nicolas.blavet@ceitec.muni.cz</p>
<p style="font-size: 0.9em; color: #999;">Report generated with the enrichment_and_gsea pipeline | `r format(Sys.time(), "%Y-%m-%d %H:%M")`</p>
</div>

</div> <!-- End container-main -->
