---
title: Gene enrichment analysis
author:
  - Bioinformatics Core Facility, Centre for Molecular Medicine, CEITEC Masaryk University
  - ^1^CEITEC-Central European Institute of Technology, Masaryk University, Kamenice 5, 625 00 Brno, Czech Republic
  - Nicolas Blavet^1^ [nicolas.blavet@ceitec.muni.cz](mailto:nicolas.blavet@ceitec.muni.cz)

output:
  html_document:
    highlight: tango

params:
  config: snakemake.params.config
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r load-packages, include=FALSE}
library(data.table)
library(jsonlite)
```

```{r variables, results='asis', echo=FALSE}
DE_FC <- "DE_feature_count"
if(dir.exists(paste0("results","/DE_RSEM"))==T){
  var.rsem <- 1
  DE_RSEM <- "DE_RSEM"
}else{
   var.rsem <- 0
}

`%!in%` <- Negate(`%in%`)

```

# Acknowledgement

- We request co-authorship (and will assist with manuscript preparation) if we have developed novel tools, algorithms, or pipelines, participated in experiment design planning, or have contributed to a biological question addressed in a manuscript. We ask that, at a minimum, you acknowledge us in a publication to which we have contributed routine analysis without further support and discussions, data management or conversion, or data submission services.
- Core Facility Bioinformatics of CEITEC Masaryk University is gratefully acknowledged for the obtaining of the scientific data presented in this paper.

# Description
```{r, results='asis', echo=TRUE}
print(params$config)
#config_json = as.data.table(fromJSON(paste0(params$config)))

```